---
title: "Power Rankings"
pagetitle: "The Footy Forest | Power Rankings"
toc: false
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(gt)

# Read power rankings CSV (overwritten each week by the R pipeline)
power <- read_csv("data/power_rankings.csv", show_col_types = FALSE)

# Read tips to get round info
tips <- read_csv("data/tips_table_for_website.csv", show_col_types = FALSE)
round_info <- tips$seas_rnd[1]
season <- str_extract(round_info, "^\\d+")
round_num <- str_extract(round_info, "\\d+$")
```

::: {.content}

::: {.section-heading}
Current Power Rankings
:::

ELO-based power ratings entering Round `r round_num` of the `r season` season. An average team sits at 100. Ratings carry over 70% between seasons and exclude home ground advantage, which is estimated separately via venue performance effects.

::: {style="max-width: 420px;"}

```{r}
#| echo: false
#| message: false
#| warning: false

power %>%
  select(Ranking, Team, Power) %>%
  gt() %>%
  tab_header(
    title = md("**Current Power Rankings**"),
    subtitle = md(paste0("*ELO ratings entering ", season, " Round ", round_num, "*"))
  ) %>%
  data_color(
    columns = c("Power"),
    palette = c("#D94F4F", "#F5F5F5", "#3A8A3A"),
    domain = c(min(power$Power), max(power$Power))
  ) %>%
  cols_align(columns = c("Ranking", "Power"), align = "center") %>%
  cols_align(columns = "Team", align = "left") %>%
  cols_width(
    "Ranking" ~ px(70),
    "Team" ~ px(180),
    "Power" ~ px(80)
  ) %>%
  tab_style(
    style = cell_borders(sides = "bottom", color = "#E8E8E8", weight = px(1)),
    locations = cells_body()
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold", size = px(11), color = "#444444"),
      cell_fill(color = "#F8F8F8"),
      cell_borders(sides = "bottom", color = "#CCCCCC", weight = px(2))
    ),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", color = "#222222"),
    locations = cells_body(columns = "Ranking")
  ) %>%
  tab_options(
    table.font.size = px(13),
    table.font.names = c("system-ui", "-apple-system", "Segoe UI", "Roboto", "sans-serif"),
    heading.title.font.size = px(18),
    heading.subtitle.font.size = px(12),
    heading.border.bottom.color = "#CCCCCC",
    table.border.top.color = "#CCCCCC",
    table.border.bottom.color = "#CCCCCC",
    table_body.border.bottom.color = "#CCCCCC",
    column_labels.padding = px(8),
    data_row.padding = px(6),
    table.width = px(350)
  )
```

:::

::: {.section-heading}
About the ratings
:::

Power ratings are calculated using a customised [ELO system](https://en.wikipedia.org/wiki/Elo_rating_system), adapted for AFL:

- **Margin capping**: Wins/losses are capped at 80 points to limit the influence of blowouts.
- **Season regression**: Ratings regress 30% toward 1000 at the start of each season.
- **No built-in HGA**: Home ground advantage is set to zero in the ELO formula. Instead, venue-specific effects are estimated separately using 5 years of performance data relative to ELO expectations.
- **K-factor of 60**: Ratings update relatively quickly in response to results.

These ratings feed into the ensemble ML model as one of four predictor variables (alongside venue effect, team player ratings, and travel differential).

:::
